Testing the motor and cognitive foundations of Paleolithic social transmission

### 2021-08-14 manuscript/manuscript.Rmd
change the layout and add the discussion section

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
subtitle: 
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Justin Pargeter 1, 2
  address: 
  - 1. Department of Anthropology, New York University, New York, NY, USA; 2. Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa
  email: justin.pargeter@nyu.edu

- name: Megan Beney Kilgore 3
  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
  email: megan.elizabeth.beney@emory.edu

- name: Cheng Liu 3
  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
  email: raylc1996@outlook.com

- name: Dietrich Stout 3
  address: 3. Department of Anthropology, Emory University, Atlanta, GA, USA
  email: dwstout@emory.edu

keywords:
- Oldowan
- Stone toolmaking
- Social learning
- Individual variation
- Cognitive aptitudes
- Motor skills

#PACS: 
#- PAC1
#- superPAC

abstract: |
  Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.

bibliography: bibliography.bib
bibstyle: spmpsci
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

# Introduction {#intro}

@@

## **Individual Differences**

@@

## **Teaching, Language, and Tool Making**

*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.* [@montagu1976: 267]

@@

## **Raw materials and knapping skill**

@@

# **Materials and Methods**

@@

## **Participants**

@@

## **Study Visit**

@@

## **Individual Difference Measures**

@@

## **Stone Tool Making**

@@

### **Raw Materials**

@@

### **Experimental Conditions**

@@

**Lithic Analysis**

@@

## **Statistical Analyses**

@@

# **Results**

@@

## **Principal Component analyses**

@@

### **Flake size and shape**

@@

### **Lithic flaking performance measures**

@@

## **Do trained, untrained, and expert knappers perform differently?**

@@

## **Does training/practice time impact flaking performance?**

@@

## **Do individual differences in motor skill and psychometric measures predict flaking performance?**

@@

### **Model 1: Individual differences and quantity flaking**

@@

*Quantity flaking $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength*

@@

*Quantity flaking $\sim$ Training condition + Highest n-back level  + BEAST score + Grip strength + Fitt’s score$\cdot$Raven’s Progressive Matrix score + Training condition$\cdot$Highest n-back level + Training condition$\cdot$BEAST score + Nodule mass (as control)*

@@

### **Model 2: Individual differences and quality flaking**

@@

*Quality flaking $\sim$ Highest n-back level  + Fitt’s score + Grip strength + Fitt’s score$\cdot$BEAST score + Grip strength$\cdot$BEAST score + Grip strength$\cdot$Fitt’s score + Training condition$\cdot$Grip strength + Nodule mass (as control)*

@@

The quality flaking model shows two significant interaction effects involving grip strength, BEAST scores, and training. The interaction between grip strength and BEAST scores has a significant positive effect on careful flaking (Est. = 0.5 [0.22, 0.86]). **Figure** illustrates this interaction whereby a subject's propensity to use social information mitigates against individual differences in grip strength. Strong subjects with a low propensity to use social information tend to perform worst in terms of quality flaking. This result makes sense if one considers that our quantity flaking model showed how social information equalizes performance levels in trained and untrained subjects. \
\
The second interaction involves grip strength and training. A one unit increase in this interaction results in a significant positive effect on careful flaking (Est. = 1 [0.51, 1.47]), but only in the untrained group. **Figure** illustrates this interaction showing how training helps even out flaking performance differences between stronger and weaker individuals. Stronger individuals perform worse in terms of quality flaking when they are untrained. Trained subjects, regardless of grip strength, perform equally well in terms of quality flaking.

# **Discussion**

we have little basis other than personal experience and/or tradition [@callahan1979; @whittaker1994; @shea2015] and theoretical speculation [@stout2013; @whiten2015] from which to assess what pedagogical techniques are most effective even in WEIRD contexts. For example, no study to date has considered how variation in teacher skill [@shea2015] or social relationship to participants might impact learning under different conditions. To properly address these questions would require a major research program, including both cross-cultural comparative studies [@barrett2020] (Barrett 2020) and more naturalistic study designs. While costly, such research would produce results of broad relevance to anthropologists, biologists, psychologists, and sociologists interested in teaching and learning, well beyond any particular implications for language evolution.

# **Conclusions**

# **Acknowledgments**

# **References**

>>>>

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

# Introduction

@@

## Individual Differences

@@

## Teaching, Language, and Tool Making

“*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*”[@montagu1976: 267]

@@

## Raw materials and knapping skill

@@

# Materials and Methods

@@

## Participants

@@

## Study Visit

@@

## Individual Difference Measures

@@

## Stone Tool Making

@@

### Raw Materials

@@

### Experimental Conditions

@@

## Lithic Analysis

@@

## Statistical Analyses

@@

# Results

@@

## Principal Component analyses

@@

### Detached Piece size and shape

@@

### Lithic flaking performance measures

@@

## Relationships between Performance Measures

This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig. Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. R2 = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. R2 = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. R2 = 0.7), but no relation in the Trained group (Fig. Xb). A qualitatively similar trend with respect to Quality vs. Quantity (Fig. Xc) did not achieve significance (F[1, 17] = 0.6, p 0.2).

Thus, it appears that Trained participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality, whereas Untrained participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see variance Table and Figure) and generally increased productivity at the expense of Quality. Experimental artifacts illustrating these trade-offs are presented in Figure.

## Do trained, untrained, and expert knappers perform differently?

@@

## Does performance change over time?

@@

## Do individual differences in motor skill and psychometric measures predict flaking performance?

@@

### Model 1: Individual differences and quantity flaking

@@

*Quantity flaking* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength

@@

*Quantity flaking* $\sim$ Training condition + Highest n-back level + BEAST score + Grip strength + Fitt's score$\cdot$Raven's Progressive Matrix score + Training condition$\cdot$Highest n-back level + Training condition$\cdot$BEAST score + Nodule mass (as control)

@@

### Model 2: Individual differences and quality flaking

@@

*Quality flaking* $\sim$ Highest n-back level + Fitt's score + Grip strength + Fitt's score$\cdot$BEAST score + Grip strength$\cdot$BEAST score + Grip strength$\cdot$Fitt's score + Training condition$\cdot$Grip strength + Nodule mass (as control)

@@

The quality flaking model shows two significant interaction effects involving grip strength, BEAST scores, and training. The interaction between grip strength and BEAST scores has a significant positive effect on careful flaking (Est. = 0.5 [0.22, 0.86]). **Figure** illustrates this interaction whereby a subject's propensity to use social information mitigates against individual differences in grip strength. Strong subjects with a low propensity to use social information tend to perform worst in terms of quality flaking. This result makes sense if one considers that our quantity flaking model showed how social information equalizes performance levels in trained and untrained subjects.

The second interaction involves grip strength and training. A one unit increase in this interaction results in a significant positive effect on careful flaking (Est. = 1 [0.51, 1.47]), but only in the untrained group. **Figure** illustrates this interaction showing how training helps even out flaking performance differences between stronger and weaker individuals. Stronger individuals perform worse in terms of quality flaking when they are untrained. Trained subjects, regardless of grip strength, perform equally well in terms of quality flaking.

### Model 3: Individual differences and quality flaking

## Behavioral observations

We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on the effect of individual differences and training on knapping outcomes. We thus focused on collecting quantitative psychometric and lithic data. However, we also considered that quantifying participant knapping behaviors as well as products could be important for future studies. To support methods development in this regard, we made ad hoc notes on observed behaviors during the experiments and video-recorded all experiments to enable later, more systematic analyses yet to be completed. However, even casual behavior observation was sufficient to reveal an unexpected effect. Whereas all trained participants copied the general posture and technique of the expert (free hand knapping seated in a chair) fully half (6) of the uninstructed participants experimented with or even knapped all of their cores using the floor as a support (Figure). Three of these participants were in the same session, which is also the only group composed of just three individuals. In this group, knapping on the ground appears to have been transmitted from one participant to the other two based on appearance order and the point of gaze of participants.

# Discussion

The most salient finding of our exploratory study is that the presence/absence of teaching clearly impacted knapping performance but did so in nuanced and individually variable ways that have not been explored in previous studies. In fact, we did not observe any significant differences in mean performance between our experimental conditions. Some non-significant tendencies toward enhanced Instructed group performance suggest that a larger participant sample might detect significant effects, but also that the size of any such effects would likely remain small. This could reasonably lead to the conclusion that teaching does not substantially facilitate early stage knapping skill acquisition [cf. @putt2014; @ohnuma1997]. Looking closer, however, we found a number of important teaching effects.

## Variance Reduction

In our experiment, the strongest effects of teaching were to reduce variance (Figure, Table) rather than shift mean values. In particular, teaching acted as a "safety net" that homogenized performance by reducing the frequency of extremely poor outcomes (i.e., learning failures). This finding provides additional support for the hypothesis that teaching would have increased the reliability of Oldowan skill reproduction [@morgan2015] while simultaneously corroborating the view that basic flaking competence can be achieved in its absence [@tennie2017]. Our results thus do not imply that teaching was required or even present during Oldowan times (but see @gärdenfors2017), but rather serve to reinforce the plausibility of co-evolutionary scenarios positing the cost/reliability of technological skill acquisition as a selection pressure favoring the evolution of teaching and language [@stout2010; @morgan2015; @stout2017].

## Knapping Behaviors

The current study complements the transmission chain design of Morgan et al. [-@morgan2015] by finding similar effects in more naturalistic learning contexts. Our design further allowed us to examine individual variation to better understand how teaching produces its effects. Whereas transmission chains are optimized to investigate iterative learning effects [but see @caldwell2020], they necessarily involve a different instructor/model for each participant. We sacrificed this iterative component in order to consider how the presence/absence of teaching affected the behavior of individuals under otherwise standardized learning conditions.

We found that a key impact of teaching was to alter basic flake production strategies, as reflected in the relationship between Total Productivity and detached piece Quality (Fig.Xb). Whereas Untrained participants achieved greater Productivity at the expense of Quality and vice versa, these dimensions were unrelated across Trained participants. Thus, even though Untrained participants achieved the highest values on each metric, only trained individuals managed to maximize both simultaneously. Indeed, a core function of teaching is to reduce the search space that learners must explore and increase the likelihood of discovering globally as opposed to locally optimum solutions [cf. @hinton1996; @stout2013]. In our study, Untrained individuals explored a greater range of basic behavioral variations not seen in the Trained group, including knapping on the floor, concentrating on working just a few cores (2-4, Figure) over the practice period, and showing less constrained nodule size preferences (Figure). It is notable that this variation occurred even in the presence of an observable expert example, suggesting it may be interesting for future experiments to address the impact of social context, expectations, and relationships on observational learning strategies [@kendal2018].

We also found a strong positive effect of Grip Strength on Total Productivity independent of learning condition (Figure). While it is tempting to interpret this with respect to the demands for hand strength specifically, it is important to remember that grip strength is strongly correlated with total muscle strength [@wind2010] and overall fitness [@sasaki2007]. Thus, it is best taken to indicate some importance of fitness generally in increasing the rate and intensity of core reduction by naïve knappers, potentially affecting rate of work and the kinetic energy of the swing as well as the handling of core and hammerstone. It thus provides further support for hypotheses positing stone tool making as a selection pressure on the functional anatomy of hand, arm, and shoulder [e.g., @williams-hatala2018], but initially appears orthogonal to variations in learning condition and knapping behaviors in our study. However, we also found that the effect of Grip Strength on per-core knapping Quantity is dependent on teaching (Figure). The absence of this effect in the uninstructed group reflects the weaker association between Total Productivity and per-core Quality across these participants (Fig.Xa) and shows Grip Strength increased uninstructed Productivity specifically by allowing them to knap more cores rather than to reduce individual cores more heavily. In keeping with this, uninstructed Grip Strength is positively correlated with Total Cores knapped (r2 = 0.54, p = 0.01). Conversely, strength allowed instructed participants to increase their average Quantity per core without affecting the total number of cores knapped (r2 = 0.18, p = 0.165). Thus, strength appears to have achieved its effects on core reduction rate and intensity in different ways, depending on teaching. This difference is likely related to the homogenizing effect of teaching on knapping rate (all instructed participants knapped 6 or more cores) and methods. Subjectively, knapping behaviors of uninstructed participants often appeared more physically demanding (e.g., greater number of non-productive blows, rapid and unregulated battering) which would imply different demands on both strength and aerobic fitness [@williams-hatala2021; @mateos2019]. However, this remains to be systematically investigated.

In this respect, it is also important to note that we do not know how well the knapping objectives and strategies communicated by the expert in our experiment correspond to actual Oldowan goals and behaviors. The instructor has successfully replicated assemblage-level patterning at Gona [@stout2019] but Oldowan behavior is variable across space and time [e.g., @braun2019] and alternative knapping methods might maximize different values (productivity, quality, effort), especially in novices [@putt2015] (Putt 2015). Nevertheless, the effect of instruction to constrain behavioral exploration and homogenize outcomes is clear. We expect that this effect would generalize to the teaching of alternative knapping goals and behaviors, although this remains to be tested.

## Learning Strategies

One major goal of this experiment was to test the viability of a moderate, two-hour, learning period for studies of skill acquisition. Unfortunately, we found that this duration was insufficient to capture learning effects for Oldowan-like flake production. The lack of performance change over the period (Figure) cannot be attributed to a ceiling effect (i.e., rapid task mastery at the outset of the practice period) as participants remained well below expert levels and continued to display the high within-individual variability typical of naïve/novice knapping [@eren2011; @pargeter2019]. This negative result was unexpected but is broadly consistent with evidence that Early Stone Age flaking, while conceptually simple, requires substantial practice for perceptual-motor skill development [@nonaka2010; @stout2015; @pargeter2020]. Future investigations of learning variation across individuals and/or experimental conditions may thus need to incorporate longer practice periods to capture skill acquisition processes. In theory, much shorter knapping trials might be used to assess the variation in initial performance across individuals and under different conditions that is captured in our study. However, the presence of substantial core-to-core variation within individuals cautions against overly brief experiments that might not provide a representative sample. Greater durations also allow for the expression of different learning strategies over time, even in the absence of directional performance change.

At a basic level, learners of any new task must balance investment in task exploration vs. exploitation of knowledge and skills already in hand [@sutton2018]. Premature exploitation risks settling for a sub-optimal local solution whereas continued exploration sacrifices more immediate payoffs. Managing this trade-off is especially challenging for complex, real-world tasks like stone knapping, and is thought to depend on the interplay of uncertainty and reward expectation [@wilson2021]. Teaching and social learning generally have the potential to provide low-cost information about task structure and payoffs [@rendell2010; @kendal2018], which if adopted, would be expected to affect exploration/exploitation decisions. Such adoption is itself known to be influenced by individual cognitive differences, for example if higher fluid intelligence allows observers to better understand observed tasks [@vostroknutov2018] or if individuals vary in their tendency to use and value social information [@toelch2014; @molleman2019].

In our study, we did not observe any effect of fluid intelligence (RPM) on knapping outcomes but did find strong interactions of learning condition with participant visuospatial working memory and social information use tendency (Figure). As expected, uninstructed individuals with higher scores on these dimensions displayed higher Total Productivity and average per-core flaking Quantity (although the effect on n-Back on Quantity did not achieve significance). We attribute these effects to increased ability to hold relevant morphological/spatial information in mind and a tendency to benefit from observing successful strategies of others, including the expert model. In contrast, instructed individuals with higher scores tended to have lower Productivity and Quantity. We interpret this unexpected effect to an increased tendency to privilege exploratory learning behavior over exploitation. In particular, we suggest that trained participants might knap more slowly and less productively if higher working memory capacity inclined them to experiment more with morphological/spatial variables highlighted by the instructor or if a predisposition to use social information use caused them to invest greater time and effort attending to and trying out observed actions and/or instructions. These suggestions remain to be tested by further work.

Unfortunately, the training period in our current experiment was insufficient to capture learning effects and so we have no evidence of the effects of these individual differences and putative exploration/exploitation tradeoffs to the ultimate achievement of expertise. A similar negative effect of instruction on knapping outcomes during early stage learning was reported by Putt et al. [-@putt2014], and has been interpreted to reflect learners experimenting with advanced techniques before they have the perceptual-motor skill to execute them [@whiten2015; @stout2015]. Such effects might be further explored with more detailed behavioral data, as opposed to purely lithic data, and with longer learning periods.

## Limitations and Prospects

Although our exploratory study produced a number of robust results with respect to the effects of instruction and individual differences on lithic products, it is clearly limited by a small sample size, short training duration, and lack of detailed quantification of observed behaviors. These are limitations that can hopefully be addressed in future studies building on the methods and evidence presented here. For example, it is notable that our study failed to document any reliable effects on knapping Quality. Obviously, this might reflect an actual lack of such effects, but it may also indicate a need for more sensitive measures and/or increased sample size and training duration to identify subtle or delayed effects. One aspect of our attempt to balance pragmatic costs and benefits in our study was to test the efficacy of relatively limited lithic analysis. More detailed ongoing analyses of core morphology and debitage features (e.g., typology, cutting edge length, platform dimensions) may yet reveal a more reliable signal of knapping quality. Results of the Quality model in particular also seem to suffer from the uneven distribution and discrete rather than continuous nature of scores on our RPM and n-Back tests. Concerns about the sampling of variation on these dimensions could be addressed with larger samples or by pre-screening participants to ensure more even representation. Alternative psychometric tests (e.g., full rather than short version of the RPM) might also provide more sensitive and continuous measures.

Another major limitation that our study shares with all other published experiments on knapping skill acquisition is that we do not address variation in social and cultural context or in teaching style. Currently, we have little basis other than personal experience/tradition [@callahan1979; @whittaker1994; @shea2015] and theoretical speculation [@stout2013; @whiten2015] from which to assess which pedagogical techniques are most effective even in WEIRD contexts. No study to date has considered how variation in teacher skill [@shea2015] or social relationship to participants might impact learning under different conditions. To properly address these questions would require a major research program, including both cross-cultural comparative studies [@barrett2020] and more naturalistic study designs. While costly, such research would produce results of broad relevance to anthropologists, biologists, psychologists, and sociologists interested in teaching and learning.

# Conclusions

# Acknowledgments

# References
